---
title: "Easy tracking analyses in R"
format: html
execute:
  eval: false
---

Once we've loaded and plotted our data, it's time to do some basic analyses. 

### Altitude over time
Nothing too fancy, just things like plotting the altitude of our tracked animal over time. Let's get into it!

There are two options to plot up your data.
One is a good ole-fashioned ggplot. If you want a static map, that's fine. The other option is a dygraph, which is an interactive plot. Dygraphs are perfect for long time-series data and data exploration, where you want to zoom in and out and inspect every nook and cranny of your dataset.

```{r}
# altitude over time - in a dygraph format
(duck1_y2_altitude_dygraph <- duck1_y2 %>%
  select(time, altitude) %>%
  xts::xts(order.by = .$time) %>%
  dygraphs::dygraph() %>%
  dygraphs::dyAxis("y", label = "Altitude (m)") %>%
  dygraphs::dyOptions(colors = "grey") %>%
  dygraphs::dyRangeSelector())

# altitude over time - ggplot
(duck1_y2_altitude_plot <- ggplot(duck1_y2, aes(x = time, y = altitude)) +
  geom_line() +
  labs(title = "Duck 1 Year 2 Altitude Over Time") +
  xlab("Time") +
  ylab("Altitude (m)") +
  theme_minimal())
```

From these graphs, you might see clear anomalous points that were missed in our previous filtering. E.g., we might want to remove points over 5000m or 10,000m
``` {r}
# we can see some points where the altitude is super high - over 5000m or even 10,000m.
# this is probably an error, so we will filter out any points over 5000m
duck1_y2 <- duck1_y2 %>%
  filter(altitude < 1000)

# then you can re-run the plots above and save them if you want

# save plot
ggsave(
  filename = "03_plots/duck1_altitude.png",
  plot = duck1_altitude_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

# Distance and time between points
We might also want to calculate the distance and time between subsequent GPS fixes. Here's the code to do it by calculating the mean and cumulative distances between points.
``` {r}
# mean distance between points - distHaversine from geosphere package
duck1_y2 <- duck1_y2 %>%
  mutate(
    distance = distHaversine(
      cbind(longitude, latitude), # find the Haversine distance between a given lat lon point
      lag(cbind(longitude, latitude)) # and the previous lat lon point (lag function shifts data back 1)
    )
  )

# check it worked
glimpse(duck1_y2)

summary(duck1_y2$distance)
#  Min.  1st Qu.   Median     Mean  3rd Qu.     Max.     NA's
#   0.00    10.34    23.11   181.32    63.81 10626.82        1

hist(
  duck1_y2$distance,
  breaks = 20,
  main = "Histogram of distances between points",
  xlab = "Distance (m)"
)

# calculate cumulative distance between each point and previous point
# it needs to account for the NA in the first row
duck1_y2 <- duck1_y2 %>%
  mutate(cumulative_distance = cumsum(replace_na(distance, 0)))
# find total cumulative distance by calling the final row
duck1_y2$cumulative_distance[nrow(duck1_y2)]
# 998852.6 m
998852.6 / 1000
# 998.8526 km

``` 

Another option is to calculate the length of the straight-line distance between our GPS points.
``` {r}
# figuring out the length of the straight-line distance between points
# (has it's issues if you have big time gaps between fixes)
# Create an sf POINT object from the GPS points (assuming WGS84)
duck1_y2_points <- st_as_sf(
  duck1_y2,
  coords = c("longitude", "latitude"),
  crs = 4326 # WGS84
)

# Example: Transform to CRS 7855 (Australia)
duck1_y2_GPS_points_7855 <- st_transform(duck1_y2_points, crs = 7855)

# Combine the points into a LINESTRING
duck1_y2_line_7855 <- duck1_y2_GPS_points_7855 %>%
  summarise(geometry = st_combine(geometry)) %>% # Combine all points into one multilinestring
  st_cast("LINESTRING") # Cast to LINESTRING

plot(duck1_y2_line_7855)

# Save to shapefile
st_write(
  duck1_y2_line_7855,
  "01_tidy_data/duck1_y2_movement_path.shp",
  delete_layer = TRUE
)

# save as RData with duck1_y2_GPS_points_7855

save(
  duck1_y2_GPS_points_7855,
  duck1_y2_line_7855,
  file = '01_tidy_data/duck1_y2_GPS_points_line_sf_7855.RData'
)
# reload sf version of duck1_y2 GPS points in 7855 projection with a line shapefile too
load('01_tidy_data/duck1_y2_GPS_points_line_sf_7855.RData')

# Calculate total movement path length (planar, no terrain)
# This is an ESTIMATE based on the length of the lines between the GPS data.
# In reality, the distance will be even larger, although we cannot say by how much.
st_length(duck1_y2_line_7855)
# 986873.7 [m]
# in km
st_length(duck1_y2_line_7855) / 1000
# 986.8737 km
# pretty similar to the method above with cumulative distance between points
```

Finally, here's some code to calculate the time lag in minutes and hours between each GPS point.
``` {r}
# calculate time lag in mins and hours between each gps point
duck1_y2 <- duck1_y2 %>%
  mutate(
    time_diff_h = as.numeric(difftime(time, lag(time), units = "hours")),
    time_diff_m = as.numeric(difftime(time, lag(time), units = "mins"))
  )
# check it worked
glimpse(duck1_y2)
# success
```

Well done! 