---
title: "move2 package: Net-Squared Displacement & more"
format: html
execute:
  eval: false
---

# Net-Squared Displacement 
Now we've done some more basic analyses, it's time to level up. When studying animal movement, net-squared displacement (NSD) is sometimes pretty useful. This is especially so when we want to learn about an animal's relationship to the places it inhabits, or at least a given point.

NSD simply measures the distance (or "displacement") from a given point (normally the first point in the dataset analysed).
From this, we can answer the following questions
- What movement mode does the animal show? Does it migrate, have a set home range, or is it nomadic?
- Has the animal settled into a set home range?
- Is the animal exploring new areas further from home/the release point? When? How long?

These are especially pertinent questions when dealing with animal translocations or reintroductions, and/or when studying migratory animals.

Don't just think about this in a 2D sense, think about altitude as well. An animal might move seasonally up to higher altitudes, which aligns well with the ways of interpreting NSD curves to determine movement modes.

# The move2 package
An easy way to calculate NSD is offered by the move2 package. Building upon the popularity and success of the original move package, this is a great resource. Check out the move2 vignettes and website to learn about what it can do!

First thing's first, we need to convert our data from a tibble or dataframe into a move2 object
```{r}
# convert data to move2 object
duck1_y2_mv <- mt_as_move2(
  duck1_y2,
  coords = c("longitude", "latitude"),
  time_column = "time",
  track_id_column = "id"
)

# inspect move2 object
duck1_y2_mv

# A <move2> with `track_id_column` "UUID" and `time_column` "time"
# Containing 1 track lasting 365 days in a
# Simple feature collection with 5668 features and 18 fields
# Geometry type: POINT
# Dimension:     XY
# Bounding box:  xmin: 144.4083 ymin: -38.26497 xmax: 144.5623 ymax: -38.1992
# CRS:           NA
# # A tibble: 5,668 × 19
#    UUID   time                     t altitude altitude_ellipsoid speed
#  * <chr>  <dttm>               <dbl>    <dbl>              <dbl> <dbl>
#  1 c4c9a… 2021-01-20 14:58:49 1.61e9      2.1               -1.6   0.3
#  2 c4c9a… 2021-01-20 15:58:16 1.61e9     11.3                7.6   0
#  3 c4c9a… 2021-01-20 17:58:49 1.61e9      0.6               -3.1   0.9
#  4 c4c9a… 2021-01-20 18:59:09 1.61e9    103.                99.2   0
#  5 c4c9a… 2021-01-20 19:58:10 1.61e9     17.1               13.4   0.3
#  6 c4c9a… 2021-01-20 20:58:38 1.61e9     48.6               44.9   0.1
#  7 c4c9a… 2021-01-20 21:58:30 1.61e9     54.7               51     0.1
#  8 c4c9a… 2021-01-20 22:58:31 1.61e9     85                 81.3   0.6
#  9 c4c9a… 2021-01-20 23:59:16 1.61e9     76.9               73.2   0.1
# 10 c4c9a… 2021-01-21 00:58:43 1.61e9     56.1               52.4   0
# # ℹ 5,658 more rows
# # ℹ 13 more variables: north_speed <dbl>, east_speed <dbl>,
# #   course <dbl>, satellite_used <dbl>, positioning_mode <dbl>,
# #   hor_accuracy <dbl>, ver_accuracy <dbl>,
# #   gps_time_consumption <dbl>, data_source <chr>, hdop <dbl>,
# #   vdop <dbl>, distance <dbl>, geometry <POINT>
# # ℹ Use `print(n = ...)` to see more rows
# Track features:
#           UUID
# 1 c4c9ad1ddc4e
```

# Data preparation and checks in move2
We can also do some other checks on the data. For example, we can check the Coordinate Reference System (CRS), which all geospatial data should be assigned. 

In case you hadn't already ordered your tracking data temporally, you can check that using mt_is_time_ordered(). You can also ensure any duplicated rows are removed. We don't want them messing up our analyses!

``` {r}
# check our CRS
sf::st_crs(duck1_y2_mv)
# currently NA
# provide an appropriate CRS for our data - EPSG:4326
sf::st_crs(duck1_y2_mv) <- 4326

# check the data is ordered by time
assertthat::assert_that(mt_is_time_ordered(duck1_y2_mv))
# check there are no duplicate rows
duck1_y2_mv <- mt_filter_unique(duck1_y2_mv)
```

# (Net-Squared) Displacement
The time has come to make our net-squared displacement columns.
I just copied the code provided on the move2 vignettes. Again, check them out!

First we will make a displacement column, then we square it. The reason for squaring displacement is basically since it becomes easier to spot changes in the data.

``` {r}
# create a displacement column
duck1_y2_mv <- duck1_y2_mv %>%
  mutate(
    displacement = c(sf::st_distance(
      !!!syms(attr(., "sf_column")),
      (!!!syms(attr(., "sf_column")))[row_number() == 1]
    ))
  )

# now make net squared displacement (square displacement)
duck1_y2_mv <- duck1_y2_mv %>%
  mutate(nsd = displacement^2)

# summary stats (in km - divided by 1000)
# displacement - km
summary(duck1_y2_mv$displacement) / 1000
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 0.000   1.468   3.577   3.212   3.708   7.507

# nsd - km2
summary(duck1_y2_mv$nsd) / 1000000
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#   0.000   2.155  12.793  13.144  13.748  56.350
```

# Plotting NSD in ggplot
Of course, no good analysis would be complete without a plot. That's the power of NSD. 

We can use a classic ggplot, like below.
```{r}
# plot nsd over time using ggplot
duck1_y2_mv %>%
  ggplot() +
  geom_line(aes(
    x = time,
    y = as.numeric(units::set_units(nsd, "km2"))
  )) +
  ylab("Distance from start point (km2)") +
  ggtitle("Duck 1 Year 2 Net Squared Displacement") +
  theme_bw()
```

# Plotting NSD using dygraphs
However! We can also consider using the dygraphs package, which is especially awesome for long time-series datasets where we might want to zoom in and out on different sections of the data.

To use the dygraphs package, we need to turn our data into a timeseries using the xts package. However, xts won't accept a move2 object; the workaround is to extract the nsd column from our move2 object and stick it back onto our original tracking tibble. 
Since move2 uses the units package, our nsd column in the move2 object is a "units" object, so we also convert it to a plain ol' numeric data type so we can successfully stick it onto our tibble.
We will also divide nsd by 1,000,000 so it is represented in km2, not m2.
``` {r}
duck1_y2 <- duck1_y2 %>%
  mutate(nsd = as.numeric(duck1_y2_mv$nsd) / 1000000) # convert from m2 to km2

# now we can make the dygraph
(duck1_y2_nsd_dygraph <- duck1_y2 %>%
  select(time, nsd) %>%
  xts::xts(x = .$nsd, order.by = .$time) %>%
  dygraphs::dygraph() %>%
  dygraphs::dyAxis("y", label = "Distance from start point (km2)") %>%
  dygraphs::dyOptions(colors = "blue") %>%
  dygraphs::dyRangeSelector())
```